Up to 25\% of beneficial mutations in protein sequences are not adaptive innovations in mammals

### 2022-11-29 manuscript/main.tex
Refactored manuscript

\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\textrm{Var}}
\newcommand{\der}{\textrm{d}}
\newcommand{\e}{\textrm{e}}
\newcommand{\avg}[1]{\left< #1 \right>} % for average
\newcommand{\Ne}{N_{\textrm{e}}}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\textrm{fix}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\Sphy}{S_{0}}
\newcommand{\Sphyclass}{\mathcal{C}}
\newcommand{\SphyMean}{\overline{\Sphy}}
\newcommand{\divStrongDel}{\Sphy < -3}
\newcommand{\divDel}{-3 < \Sphy < -1}
\newcommand{\divWeakDel}{-1 < \Sphy < 0}
\newcommand{\divWeakAdv}{0 < \Sphy < 1}
\newcommand{\divAdv}{ \Sphy > 1}
\newcommand{\PdivStrongDel}{\proba \left[ \divStrongDel \right]}
\newcommand{\PdivDel}{\proba \left[ \divDel \right]}
\newcommand{\PdivWeakDel}{\proba \left[ \divWeakDel \right]}
\newcommand{\PdivWeakAdv}{\proba \left[ \divWeakAdv \right]}
\newcommand{\PdivAdv}{\proba \left[ \divAdv \right]}
\newcommand{\given}{\mid}
\newcommand{\Spop}{S}
\newcommand{\SpopMean}{\overline{\Spop}}
\newcommand{\polyDel}{\Spop < -1}
\newcommand{\polyNeutral}{-1 < \Spop < 1}
\newcommand{\polyAdv}{ \Spop > 1}
\newcommand{\PpolyDel}{\proba \left[ \polyDel \right]}
\newcommand{\PpolyNeutral}{\proba \left[ \polyNeutral \right]}
\newcommand{\PpolyAdv}{\proba \left[ \polyAdv \right]}

\renewcommand{\baselinestretch}{1.5}
\linenumbers

@@

@@

    \begin{abstract}
        Mutations can be beneficial by bringing a new innovation to their bearer, allowing them to adapt to change in environments or in selection.
        However, mutations can also be beneficial because they are repairing previous deleterious changes, simply restoring existing functions
        In this study, we first estimated selective effects of mutations inside protein coding sequence, under a model assuming no adaptation at the phylogenetic scale.
        We then estimated the proportion of beneficial mutations that are not adaptive innovations, and subsequently estimated their proportion among all beneficial mutations at the population scale.
        Our work confirms that deleterious substitutions have accumulated in mammals and are currently being eliminated, resulting in up to 25\% of beneficial mutations that are not adaptive innovations, but instead are repairing previous deleterious changes.

    \end{abstract}

    \keywords{Adaptation \and beneficial mutations \and phylogenetic \and population-genetics \and codon models }

    Adaptation is considered as the main process driving the diversity of forms and functions across the tree of life, as it enables species to access new ecological niches and also respond to changes in their environment~\cite{darwin_origin_1859}.
    In this context of adaptative evolution, a change in environment is echoed as a change of species' forms and functions, ultimately leaving signatures of accelerated changes in the species' genome~\cite{merrell_adaptive_1994}.
    This signature of an accelerated rate of changes, generating an excess of beneficial mutations is used to quantify adaptation (fig.~\ref{fig:fitness-landscape}A), for specific genes along a specific lineage~\cite{mcdonald_adaptative_1991, smith_adaptive_2002a, welch_estimating_2006}.
    The current availability of large scale genomic data and the development of theoretical models allowed to detect and quantify adaption across genes and lineages, which is now a common practice in evolutionary biology~\cite{yang_statistical_2000, eyre-walker_genomic_2006, moutinho_variation_2019}.
    Although these approaches led to a better understanding of the processes shaping molecular evolution, it had the effect to merely associate beneficial mutations with adaptive evolution, which should be dissociated since adaptive evolution is not the only process which can lead to beneficial mutations~\cite{charlesworth_other_2007, mustonen_fitness_2009}.
    Indeed, even in a constant environment for which the population is already adapted, a deleterious mutation in an individual can reach fixation in the population by genetic drift~\cite{Ohta1992}.
    Subsequently, a mutation restoring the ancestral state will thus be beneficial (fig.~\ref{fig:fitness-landscape}B), even though no change in the environment occurred~\cite{hartl_compensatory_1996, sella_application_2005, mustonen_fitness_2009, cvijovic_fate_2015}.
    More precisely, this restoration of the ancestral state can happen at another loci, in which case it is referred as compensating mutations~\cite{hartl_compensatory_1996, mustonen_fitness_2009}.
    Conversely, the restoration of the ancestral state can happen at the same locus, in which case it is referred as a beneficial back-mutation~\cite{piganeau_estimating_2003, charlesworth_other_2007}, which are the focus of this study.
    Taken together, not all beneficial mutations contribute to genetic innovation, and truly adaptive evolution can only be well estimated under the background of beneficial back-mutations~\cite{keightley_what_2010, rice_evolutionarily_2015}.
    It is thus essential to distinguish the underlying reason for the advantage brought by a beneficial mutation, a back-mutation repairing damaged genomes or creating an adaptive innovations.

    For an individual carrying a beneficial mutation, whether this mutation is new innovation as response to a change in environment or a beneficial back-mutation is indistinguishable.
    Likewise, at the level of the population, both will result in a positive transmission bias of the beneficial allele.
    However, their consequences at the macro-evolutionary scale are fundamentally different.
    While adaptive evolution promotes phenotype diversification (fig.~\ref{fig:fitness-landscape}C), beneficial back-mutations promotes phenotype stability and preserves well established biological systems (fig.~\ref{fig:fitness-landscape}D).
    Fundamentally, adaptive evolution is unpredictable because it caused by an unpredictable change in the environment and the underlying fitness landscape~\cite{bazykin_changing_2015}.
    On the other hand, beneficial back-mutations are predictable if the underlying stable fitness landscape is known.
    As a result, a way to quantify the amount of back-mutation is to estimate the underlying fitness landscape, assuming it is constant and stable across time.

    In this study, fitness landscapes of protein-coding genes are estimated from multi-species sequence alignments.
    Multi-species sequence alignments can be used as a proxy for fitness, since for example a deleterious amino acid will be more likely to be absent in the alignment.
    One of the most used methods to make these predictions is SIFT (Sort Intolerant From Tolerant), which reliably detects deleterious mutations in many taxa~\cite{ng_sift_2003, vaser_sift_2016}.
    However, SIFT scores is not interpretable as a selection coefficient.
    Second, we know that a lot of mechanisms can influence amino acid frequency other than fitness (phylogenetic inertia, mutation biases, GC-biased gene conversion, mutational paths) which are not accounted for in SIFT\@.
    Contrarily, so-called mutation-selection codon models are rooted within a population-genetic formalism~\cite{halpern_evolutionary_1998, mccandlish_modeling_2014}.
    These phylogenetic codon models can disentangle processes of mutation at the DNA level and selection at the amino acid level.
    Such codon models provide a nearly-neutral framework of evolution by estimating the fitness landscape over amino acid sequences for each site of the sequence (fig~\ref{fig:method}A) while taking into account non-adaptive processes impacting codon frequency~\cite{halpern_evolutionary_1998, rodrigue_mechanistic_2010, tamuri_estimating_2012, latrille_improved_2022a}.
    From the fitness landscape obtained by fitting the model to the data, one can thus compute the scaled selection coefficient, denoted $\Sphy$, of any non-synonymous mutations.
    Importantly, mutation-selection codon models assume that the underlying fitness landscape is stable along the phylogeny (as in fig.~\ref{fig:fitness-landscape}B), hence the null subscript $0$ for $\Sphy$.
    As a result, a mutation with a positive $\Sphy$ would be driving the current site toward a fitter amino-acid, and thus considered a beneficial back-mutation repairing existing functions.
    Altogether, our expectation if that beneficial back-mutations are detectable and scattered across the genomes, while at the same time some slightly deleterious mutations are reaching fixations at other loci, generating a balance at which genomes are constantly both damaged and repaired at different loci.
    Even though each of these back-mutations is expected to have a small beneficial effect on their bearer, and that our assumptions may not always be valid, we expect that the genome-wide signature of beneficial back-mutations can be detected and quantified.

    By integrating datasets at different scales (phylogeny and population), we can predict whether substitutions in terminal lineages or any mutation currently segregating in a population are beneficial back-mutations, i.e with a positive $\Sphy$ (fig~\ref{fig:method}A\&B).
    Such a dataset can allow us to address different questions regarding the selection affecting protein-coding DNA sequences.
    Are we empirically observing beneficial back-mutations that reached fixation recently, and one that are currently segregating in populations?
    Are they effectively beneficial for the individual carrying it?
    Conversely, what is the probability for a beneficial mutation to be a back-mutation?
    In other words, can we find empirical evidence and can we quantify the amount of beneficial mutations that are restoring damaged genomes instead of creating adaptive innovations.
    To answer these questions, we estimated the fitness landscape of amino-acids for long-term evolution using available multiple sequence alignments between mammalian species~\cite{ranwez_orthomam_2007, howe_ensembl_2021}.
    We then integrated data at the phylogenetic and population scale to estimate the amount of beneficial mutations across the entire exome (fig.~\ref{fig:method}A-F), for 28 populations from 6 genera (\textit{Equus},  \textit{Bos}, \textit{Capra}, \textit{Ovis}, \textit{Chlorocebus} and \textit{Homo}).

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.fitness-landscape}
        \caption{
            Panel A and B: at a given site inside a protein-coding DNA sequence, the different amino acids (x-axis) have different fitnesses (y-axis).
            Under a fluctuating fitness landscape (panel A), these fitnesses are changing with time.
            The protein sequence is always lagging behind the moving target defined by the amino acid fitnesses, and since substitutions are accepted preferentially if they are in the direction of this target, substitutions are on average adaptive.
            At the phylogenetic scale (panel C), this phenomena promotes phenotype diversification across species.
            Under a stable fitness landscape (panel B), all the mutations reaching fixation are either slightly deleterious and reaching fixation due to drift, or are beneficial back-mutations restoring a more optimal amino acid.
            Scarce beneficial back-mutations are likely to reach fixation and a bulk of deleterious mutations are eventually reaching fixation due to their sheer mass.
            Across the genome, this process generate a balance at which genomes are constantly both damaged and repaired at different loci.
            At the phylogenetic scale (panel D), this phenomena promotes promotes phenotype stability and preserves well established biological systems.
        }
        \label{fig:fitness-landscape}
    \end{figure*}

@@

    \subsection*{Back-mutations in the terminal lineages}
    We first probed for empirical evidence of beneficial back-mutation in the terminal lineages, which should be effectively beneficial.
    We thus reconstructed the substitutions occurring along the terminal branches of the phylogenetic tree.
    For each substitution, we estimated its scaled selection coefficient ($\Sphy$) based on the stable fitness landscape of amino-acids estimated at the scale of the mammals' phylogeny (fig.~\ref{fig:method}A).
    We showed that among all the substitutions in the terminal lineages, between 10 and 13\% had $\Sphy > 1$ (table~S1) and between 25 and 30\% had $\Sphy > 0$ (table~S2).
    Substitutions with $\Sphy > 1$ have been beneficial since the ratio of non-synonymous over synonymous divergence, called $\dnds$, was estimated to be between 1.15 and 1.68 in the different lineages (table~S1).
    This meant that beneficial back-mutations with $\Sphy > 1$ reach fixation more frequently than synonymous mutations, which are supposed to be neutral, thus are effectively beneficial.
    This also means that the ratio $\dnds$ usually used to detect adaptation while relying on substitutions between close species~\cite{mcdonald_adaptative_1991, galtier_adaptive_2016} or at the phylogenetic scale~\cite{goldman_codonbased_1994, yang_codonsubstitution_2002} is overinflated due to beneficial back-mutations.
    Then, $\dnds$ computed exclusively on non-synonymous sites with a $\Sphy < 0$ only represent the rate of evolution discarding beneficial back-mutations by construction.
    Thus, if we divide the $\dnds$ estimated on all substitutions by the one estimated only on substitutions predicted to be deleterious, we directly get the proportion of the rate of evolution which is contributed by beneficial back-mutations (see methods).
    This way, we estimated that between 22\% and 27\% of $\dnds$ is inflated because of beneficial back-mutations (table~S2).

    \subsection*{Back-mutations in the populations}
    Given that beneficial back-mutations are effectively selected for, we then ask which proportion of new mutations are beneficial back-mutations and are also beneficial at the population scale (fig.~\ref{fig:method}C-E).
    For each population with available polymorphism data, we first reconstructed the ancestral sequence (instead of the reference sequence) at the base of the population genealogy.
    From this reference sequence, we computed the scaled selection coefficient ($\Sphy$) for each possible non-synonymous mutations.
    We thus obtained the distribution of fitness effects (DFE) of new mutations from the ancestral sequence in \textit{Homo sapiens} (fig.~\ref{fig:homo-afr-results}A), under the assumption of the stable fitness landscape.
    Because $\Sphy$ is based on estimation of fitnesses at the mammalian scale, mutations with $\Sphy>1$ are considered beneficial back-mutations.
    Beneficial back-mutations only represents ~1.5\% of all possible new mutations (red in fig.~\ref{fig:homo-afr-results}A, table~S3).
    However, when focusing on the mutations that are observed in populations, namely single nucleotide polymorphism (SNP), beneficial back-mutations represents ~8.5\% of all observed mutations (red in fig.~\ref{fig:homo-afr-results}B, table~S3).

    Additionally, in populations, the frequencies at which SNPs are segregating is informative of their selective effect, e.g. deleterious SNPs are segregating at lower frequencies because they are purified.
    As such, methods have been developed to estimate the distribution of fitness effect (DFE) at the population scale by gathering information across many SNPs~\cite{eyre-walker_distribution_2006a, eyre-walker_estimating_2009a, galtier_adaptive_2016, tataru_inference_2017}.
    These methods offer a unique opportunity to compare the DFE at the population scale ($\Spop$) and at the phylogenetic scale ($\Sphy$).
    In order to estimate the DFE at the population level and gather robust statistical estimates, we grouped SNPs observed at the population level (e.g.~in humans of African descent in fig.~\ref{fig:homo-afr-results}) in 5 classes of predicted selection coefficient: severely deleterious with $\divStrongDel$, deleterious with $\divDel$, weakly deleterious with $\divWeakDel$, weakly beneficial with $\divWeakAdv$ and beneficial with $\divAdv$ (fig.~\ref{fig:homo-afr-results}B).
    In each of theses classes, the number of derived alleles in the population for each SNPs is used to generate a site-frequency spectrum (SFS) (fig.~\ref{fig:homo-afr-results}C).
    Compared to synonymous mutations which are supposedly neutral (black line in fig.~\ref{fig:homo-afr-results}C), mutations predicted to be deleterious (blue and green lines in fig.~\ref{fig:homo-afr-results}C) were, as expected, underrepresented at high frequencies in the population, a signature of ongoing purifying selection.
    However, these observations remain qualitative.

    To obtain a quantitative estimate of the distribution of selection coefficient for each category of SNPs, we used the model polyDFE~\cite{tataru_inference_2017, tataru_polydfe_2020}.
    This model uses the count of derived alleles for each frequency class of the SFS to infer the distribution of fitness effects (DFE) of the region under study.
    The model uses the number of sites ($L$) for each category (fig.~\ref{fig:homo-afr-results}A, see Material and Methods), the counts of synonymous derived allele as a neutral model, which allow to control for demographic effects, and polarization errors when inferring the ancestral state.
    We could estimate the proportion of beneficial mutations ($\PpolyAdv$), nearly-neutral mutations ($\PpolyNeutral$) and deleterious mutations ($\PpolyDel$) for each class of selection previously defined (fig.~\ref{fig:homo-afr-results}D).
    We found that SNPs predicted to be deleterious based on the selection coefficient $\Sphy$ estimated across the phylogeny of mammals were effectively purged at the population scale.
    Purifying selection is therefore largely predictable and amino acids with low fitness across mammals are effectively purged away in each population.
    Then, SNPs predicted to be beneficial based on the selection coefficient $\Sphy$ estimated at the scale of the phylogeny of mammals are indeed beneficial for individuals currently bearing them.
    These results confirmed that selection towards amino acids that are restoring existing function is ongoing in these populations.
    Between these two extreme cases, SNPs predicted to be weakly selected (cyan and orange in fig.~\ref{fig:homo-afr-results}) were effectively composed of neutral, beneficial and deleterious mutations.

>>>>

\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\textrm{Var}}
\newcommand{\der}{\textrm{d}}
\newcommand{\e}{\textrm{e}}
\newcommand{\avg}[1]{\left< #1 \right>} % for average
\newcommand{\Ne}{N_{\textrm{e}}}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\textrm{fix}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\Sphy}{S_{0}}
\newcommand{\Sphyclass}{\mathcal{C}}
\newcommand{\SphyMean}{\overline{\Sphy}}
\newcommand{\divStrongDel}{\Sphy < -3}
\newcommand{\divDel}{-3 < \Sphy < -1}
\newcommand{\divWeakDel}{-1 < \Sphy < 0}
\newcommand{\divWeakAdv}{0 < \Sphy < 1}
\newcommand{\divAdv}{ \Sphy > 1}
\newcommand{\PdivStrongDel}{\proba \left[ \divStrongDel \right]}
\newcommand{\PdivDel}{\proba \left[ \divDel \right]}
\newcommand{\PdivWeakDel}{\proba \left[ \divWeakDel \right]}
\newcommand{\PdivWeakAdv}{\proba \left[ \divWeakAdv \right]}
\newcommand{\PdivAdv}{\proba \left[ \divAdv \right]}
\newcommand{\given}{\mid}
\newcommand{\Spop}{S}
\newcommand{\SpopMean}{\overline{\Spop}}
\newcommand{\polyDel}{\Spop < -1}
\newcommand{\polyNeutral}{-1 < \Spop < 1}
\newcommand{\polyAdv}{ \Spop > 1}
\newcommand{\PpolyDel}{\proba \left[ \polyDel \right]}
\newcommand{\PpolyNeutral}{\proba \left[ \polyNeutral \right]}
\newcommand{\PpolyAdv}{\proba \left[ \polyAdv \right]}
\newcommand{\AdvMean}{\beta_b}
\newcommand{\DelMean}{\beta_d}
\renewcommand{\baselinestretch}{1.5}
\linenumbers

@@

% The submission checklist is available at:
% https://docs.google.com/document/d/1aHCF1on0mHTK2DoroPCLP_xto4xwHnGJCGwaL1Vm2UE/edit?usp=sharing

@@

%TC:endignore
    \begin{abstract}
        Mutations can be beneficial by bringing an innovation to their bearer, allowing them to adapt to changes in environments or in selective pressure.
        However, mutations can also be beneficial because they are repairing previous deleterious changes, simply restoring existing functions.
        In this study, we first estimated selective effects of mutations inside mammalian protein coding sequences, under a model assuming no adaptation at the phylogenetic scale.
        We then estimated the proportion of beneficial mutations that are not adaptive innovations among all beneficial mutations at the population scale.
        Our work confirms that deleterious substitutions have accumulated in mammals and are currently being eliminated.
        In modern humans, it results in around 25\% of beneficial mutations that are not adaptive innovations, but instead are repairing previous deleterious changes.
    \end{abstract}

    \keywords{Adaptation \and Back-mutation \and Phylogenetic \and Population-genetics \and Codon models }

    Adaptation is considered as the main process shaping the diversity of forms and functions across the tree of life~\cite{darwin_origin_1859}.
    It enables species to access new ecological niches and also respond to changes in their environment~\cite{darwin_origin_1859}.
    The notion of adaptation, in its evolutionary sense, is tightly linked to the notion of change, both of the environment, and of the species responding to the latter~\cite{merrell_adaptive_1994}.
    In genomes, adaptation leaves marks of accelerated evolution since beneficial mutations fix faster than neutral ones (fig.~\ref{fig:fitness-landscape}A).
    The signature of an accelerated rate of substitutions can thus be interpreted as a sign of adaptation~\cite{mcdonald_adaptative_1991, smith_adaptive_2002a, welch_estimating_2006}.
    The current availability of large scale genomic data and the development of theoretical models allowed to detect and quantify adaption across genes and lineages, which is now a common practice in evolutionary biology~\cite{yang_statistical_2000, eyre-walker_genomic_2006, moutinho_variation_2019}.
    Although these approaches led to a better understanding of the processes shaping the rates of molecular evolution, it had the effect to associate beneficial mutations with adaptive evolution, with a loss of the notion of change.
    However, we argue that they should be dissociated since adaptive evolution is not the only process which can lead to beneficial mutations~\cite{charlesworth_other_2007, mustonen_fitness_2009}.
    Indeed, even in a constant environment for which the population is already adapted, a deleterious mutation in an individual can reach fixation in the population by genetic drift~\cite{Ohta1992}.
    Subsequently, a mutation restoring the ancestral fitness will thus be beneficial (fig.~\ref{fig:fitness-landscape}B), even though no change in the environment occurred~\cite{hartl_compensatory_1996, sella_application_2005, mustonen_fitness_2009, cvijovic_fate_2015}.
    More precisely, this restoration of the ancestral fitness can happen at another loci, in which case it is referred as compensating mutations~\cite{hartl_compensatory_1996, mustonen_fitness_2009}.
    Conversely, the restoration of the ancestral fitness can happen at the same locus, in which case it is referred as a beneficial back-mutation~\cite{piganeau_estimating_2003, charlesworth_other_2007}, which are the focus of this study.
    If compensatory mutations still induce changes in the genetic sequence and thus genetic diversification, beneficial-back mutations reduce the diversity and do not contribute to genetic innovation, they cancel the deleterious changes.
    Taken together, not all beneficial mutations contribute to genetic innovation, and truly adaptive evolution can only be well estimated under the background of beneficial back-mutations~\cite{keightley_what_2010, rice_evolutionarily_2015}.
    % Philosophically, our distinction between beneficial back-mutations and adaptation can appear to be far-fetched, and one can argue that beneficial back-mutations is adaptation to its own deteriorating genome.
    % However, there is a difference between adapting to a mutation, for instance by the mean of a compensatory mutation elsewhere, and cancelling the mutation, by the mean of a substitution on the misfit amino-acid.

    For an individual carrying a beneficial mutation, whether this mutation is a new innovation responding to a change in environment or a beneficial back-mutation is indistinguishable~\cite{charlesworth_other_2007}.
    Likewise, at the level of the population, both will result in a positive transmission bias of the beneficial allele.
    However, their consequences at the macro-evolutionary scale are fundamentally different.
    While adaptive evolution promotes phenotype diversification (fig.~\ref{fig:fitness-landscape}C), beneficial back-mutations promote phenotype stability and preserves well established biological systems (fig.~\ref{fig:fitness-landscape}D).
    Fundamentally, the direction of adaptive evolution is unpredictable because it is caused by an unpredictable change in the environment and the underlying fitness landscape~\cite{bazykin_changing_2015}.
    On the other hand, beneficial back-mutations are predictable if the underlying stable fitness landscape is known.
    As a result, a way to quantify the amount of back-mutation is to estimate the underlying fitness landscape, assuming it is constant and stable across time.

    In this study, fitness landscapes of protein-coding genes are estimated from multi-species sequence alignments.
    Multi-species sequence alignments can be used to estimate amino acid fitness, since for example a deleterious amino acid will be more likely to be absent in the alignment.
    One of the most used methods to make these predictions is SIFT (Sort Intolerant From Tolerant), which reliably detects deleterious mutations in many taxa~\cite{ng_sift_2003, vaser_sift_2016}.
    However, SIFT scores is not interpretable as a selection coefficient.
    Moreover, SIFT scores do not acknowledge some key processes that can influence the abundance of an amino-acid in an alignment. For instance, it does not take into account phylogenetic inertia, mutation biases, GC-biased gene conversion, or mutational paths.
    Contrarily, so-called mutation-selection codon models are rooted within a population-genetic formalism~\cite{halpern_evolutionary_1998, mccandlish_modeling_2014}.
    These phylogenetic codon models can disentangle processes of mutation at the DNA level and selection at the amino acid level.
    Such codon models provide a nearly-neutral framework of evolution by estimating the fitness landscape over amino acid sequences for each site of the sequence while taking into account non-adaptive processes impacting codon frequency~\cite{halpern_evolutionary_1998, rodrigue_mechanistic_2010, tamuri_estimating_2012, latrille_improved_2022a}.
    Importantly, mutation-selection codon models assume that the underlying fitness landscape is stable along the phylogeny, hence is a null model of evolution without adaptation.
    From the fitness landscape obtained by fitting the model to the multi-species sequence alignments (fig~\ref{fig:method}A), one can compute the scaled selection coefficient under this null model of evolution, denoted $\Sphy$, of any non-synonymous mutations.
    In this context, a mutation with a positive $\Sphy$ would be driving the current site toward a fitter amino-acid, and thus considered a beneficial back-mutation (fig~\ref{fig:method}B).
    At the genome-wide level, our expectation is that slightly deleterious mutations that are reaching fixations due to drift are scattered across the genome, such that beneficial back-mutations concurrently segregating are also expected to be spread across genes, generating a balance at which genomes are constantly both damaged and repaired simultaneously at different loci.
    Even though each of these back-mutations is expected to have individually a small beneficial effect on their bearer, we expect that the genome-wide signature of beneficial back-mutations can be detected and quantified.

    In this study, by integrating datasets at different scales (phylogeny and population), we seek to predict whether substitutions in terminal lineages or any mutation currently segregating in a population are beneficial back-mutations, i.e with a positive $\Sphy$.
    Such a dataset can allow us to address different questions regarding the selection affecting protein-coding DNA sequences.
    Can we observe beneficial back-mutations that reached fixation recently, and one that are currently segregating in populations?
    Are they effectively beneficial for the individual carrying it?
    Conversely, what is the probability for a beneficial mutation to be a back-mutation?
    In other words, can we find empirical evidence and can we quantify the amount of beneficial mutations that are restoring damaged genomes instead of creating adaptive innovations.
    To answer these questions, we estimated the fitness landscape of amino-acids for long-term evolution using available multiple sequence alignments between mammalian species~\cite{ranwez_orthomam_2007, howe_ensembl_2021}.
    We then integrated data at the phylogenetic and population scale to estimate the amount of beneficial mutations across the entire exome (fig.~\ref{fig:method}A-F), for 28 populations from 6 genera (\textit{Equus},  \textit{Bos}, \textit{Capra}, \textit{Ovis}, \textit{Chlorocebus} and \textit{Homo}).

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=\textwidth, page=1] {artworks/figure.fitness-landscape}
        \caption{
            Panel A and B: at a given site inside a protein-coding DNA sequence, the different amino acids (x-axis) have different fitnesses (y-axis).
            Under a fluctuating fitness landscape (panel A), these fitnesses are changing with time.
            The protein sequence is always lagging behind the moving target defined by the amino acid fitnesses, and since substitutions are accepted preferentially if they are in the direction of this target, substitutions are on average adaptive.
            At the phylogenetic scale (panel C), this phenomena promotes phenotype diversification across species.
            Under a stable fitness landscape (panel B), all the mutations reaching fixation are either slightly deleterious and reaching fixation due to drift, or are beneficial back-mutations restoring a more optimal amino acid.
            At the phylogenetic scale (panel D), this phenomena promotes promotes phenotype stability and preserves well established biological systems.
            At the genome level, slightly deleterious mutations that have reached fixation are supposedly scattered at different loci, such that beneficial back-mutations should also be spread across the genome.
            We thus expect that the genome-wide signature of beneficial back-mutations can be detected and quantified, even though each mutation has individually a small beneficial effect on their bearer.
        }
        \label{fig:fitness-landscape}
    \end{figure*}

@@

    \subsection*{Back-mutations in the terminal lineages}
    We first investigated whether we could find empirical evidence of beneficial back-mutation in the terminal branches of mammals.
    We thus reconstructed the substitutions occurring along the terminal branches of the phylogenetic tree.
    For each substitution, we estimated its scaled selection coefficient ($\Sphy$) based on the stable fitness landscape of amino-acids estimated at the scale of the mammals' phylogeny (fig.~\ref{fig:method}A).
    We showed that among all the substitutions in the terminal branches, between 10 and 13\% had $\Sphy > 1$ (table~S1) and between 25 and 30\% had $\Sphy > 0$ (table~S2).
    For substitutions with $\Sphy > 1$, the ratio of non-synonymous over synonymous divergence, called $\dnds$, was estimated to be between 1.15 and 1.68 in the different lineages (table~S1).
    This meant that beneficial back-mutations with $\Sphy > 1$ reach fixation more frequently than synonymous mutations, which are supposed to be neutral, thus are effectively beneficial.
    This also means that the ratio $\dnds$ usually used to detect adaptation while relying on substitutions between close species~\cite{mcdonald_adaptative_1991, galtier_adaptive_2016} or at the phylogenetic scale~\cite{goldman_codonbased_1994, yang_codonsubstitution_2002} is overinflated due to beneficial back-mutations.
    To alleviate this bias, $\dnds$ is computed exclusively on non-synonymous sites with a $\Sphy < 0$, representing the rate of evolution discarding beneficial back-mutations by construction.
    By dividing $\dnds$ estimated on all substitutions by the one estimated only on substitutions predicted to be deleterious, we directly get the proportion of the rate of evolution which is contributed by beneficial back-mutations (see Material \& Methods).
    This way, we estimated that between 22\% and 27\% of $\dnds$ is inflated because of beneficial back-mutations while they only represent between 1\% and 2\% of mutational opportunities (table~S2).

    \subsection*{Back-mutations in the populations}
    Given that beneficial back-mutations are effectively selected for, we then ask which proportion of new mutations are beneficial back-mutations and are also beneficial at the population scale (fig.~\ref{fig:method}C-E).
    For each population with available polymorphism data, we first reconstructed the ancestral genome (instead of the reference genome) at the base of the population genealogy.
    From this reference sequence, we computed the scaled selection coefficient ($\Sphy$) for each possible non-synonymous mutations, weighted by their mutational opportunity (see Material \& Methods).
    We thus obtained the distribution of fitness effects (DFE) of new mutations from the ancestral sequence in \textit{Homo sapiens} (fig.~\ref{fig:homo-afr-results}A), under the assumption of the stable fitness landscape.
    Because $\Sphy$ is based on estimation of fitnesses at the mammalian scale, mutations with $\Sphy>1$ are considered beneficial back-mutations.
    Beneficial back-mutations only represents \~1.5\% of all possible new mutations (red in fig.~\ref{fig:homo-afr-results}A, table~S3).
    However, when focusing on the mutations that are observed in populations, namely single nucleotide polymorphism (SNP), beneficial back-mutations represents \~8.5\% of all observed mutations (red in fig.~\ref{fig:homo-afr-results}B, table~S3).

    Additionally, in populations, the frequencies at which SNPs are segregating is informative of their selective effect, e.g. deleterious SNPs are segregating at lower frequencies because they are purified.
    As such, methods have been developed to estimate the distribution of fitness effect (DFE) at the population scale by gathering information across many SNPs~\cite{eyre-walker_distribution_2006a, eyre-walker_estimating_2009a, galtier_adaptive_2016, tataru_inference_2017}.
    These methods offer a unique opportunity to compare selection coefficients estimated at the population scale ($\Spop$) and at the phylogenetic scale ($\Sphy$).
    In order to estimate the DFE at the population level and gather robust statistical estimates, we grouped SNPs observed at the population level (e.g.~in humans of African descent in fig.~\ref{fig:homo-afr-results}) in 5 classes of predicted selection coefficient: severely deleterious with $\divStrongDel$, deleterious with $\divDel$, weakly deleterious with $\divWeakDel$, weakly beneficial with $\divWeakAdv$ and beneficial with $\divAdv$ (fig.~\ref{fig:homo-afr-results}B).

    To obtain a quantitative estimate of the distribution of selection coefficient for each category of SNPs, we used the model polyDFE~\cite{tataru_inference_2017, tataru_polydfe_2020}.
    This model uses the count of derived alleles for each frequency class (fig.~\ref{fig:homo-afr-results}C) to infer the distribution of fitness effects (DFE).
    Moreover, the model relies on the mutational opportunities ($L$) for each category of $\Sphy$ (fig.~\ref{fig:homo-afr-results}A, see Material \& Methods) as well as the count of synonymous derived alleles used as a neutral model, which allow to control for demographic effects, and polarization errors when inferring the ancestral states.
    We could estimate the proportion of beneficial mutations ($\PpolyAdv$), nearly-neutral mutations ($\PpolyNeutral$) and deleterious mutations ($\PpolyDel$) for each class of selection previously defined (fig.~\ref{fig:homo-afr-results}D).
    We found that SNPs predicted to be deleterious based on the selection coefficient $\Sphy$ estimated across the phylogeny of mammals were effectively purged at the population scale (blue and green columns in fig.~\ref{fig:homo-afr-results}D)
    Purifying selection is therefore largely predictable and amino acids with low fitness across mammals are effectively purged away in each population.
    Then, SNPs predicted to be beneficial based on the selection coefficient $\Sphy$ estimated at the scale of the phylogeny of mammals are indeed beneficial for individuals currently bearing them (red column in fig.~\ref{fig:homo-afr-results}D).
    These results confirmed that selection towards amino acids that are restoring existing function is ongoing in these populations.
    Between these two extreme cases, SNPs predicted to be weakly selected (cyan and orange columns in fig.~\ref{fig:homo-afr-results}D) were effectively composed of neutral, beneficial and deleterious mutations.

    We replicated our experiment across 28 populations of different mammal species (fig.~\ref{fig:heatmap}) and found that SNPs predicted to be deleterious were indeed purged at the population scale (blue and green rows in fig.~\ref{fig:heatmap}A).
    However, SNPs predicted to be weakly beneficial ($\divWeakAdv$) were composed of a mix of neutral and selected mutations with varying proportions across the different populations (cyan and orange row in fig.~\ref{fig:heatmap}B).
    Finally, SNPs predicted to be beneficial back-mutations ($\divAdv$) were found to be beneficial with varying proportions across populations (red row in fig.~\ref{fig:heatmap}C).
    The variable proportions found between populations can be explained by the effective number of individuals in the population ($\Ne$), which is well-known to be a major driver of selection efficacy.
    Since $\Ne$ is not directly measurable, we used the Watterson $\theta_w$ as a proxy to order populations by their levels of synonymous diversity.
    Higher synonymous diversity is associated with a smaller proportion of nearly-neutral mutations (fig.~\ref{fig:diversity}A\&B).
    This result confirms the nearly-neutral theory and suggests that populations with higher diversity (e.g.~\textit{Bos} or \textit{Ovis}) are more likely to discriminate between beneficial and deleterious mutations (fig~S1-S9), while mutations in populations with low diversity (e.g.~\textit{Homo}) are effectively nearly-neutral.
